---
title: "Pair trading strategy"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyquant)
library(magrittr)
library(ggplot2)
```

```{r coke-pepsi}
# download daily prices for Coke and Pepsi stock from Yahoo Finance
coke <- tq_get("COKE", get = "stock.prices", 
               from = "2013-01-01", to = "2014-01-01")
pepsi <- tq_get("PEP", get = "stock.prices", 
                from = "2013-01-01", to = "2014-01-01")

# it is assumed that if Coke goes up on a particular day, 
# then so should Pepsi. 
# as a result, we focus on price changes of these two stocks

calculate_price_change <- function(stock_price_df) {
  price_change_df <- stock_price_df %>% 
    tq_mutate(select = adjusted, mutate_fun = diff.xts) %>% 
    dplyr::select(symbol, date, diff.xts) %>% 
    dplyr::filter(!is.na(diff.xts)) # the first entry naturally contains a missing value (NA) because there is no previous price
  
  return(price_change_df)
}


coke_price_change <- calculate_price_change(coke) 
pepsi_price_change <- calculate_price_change(pepsi)

# make a data frame with daily price change for Coke and Pepsi
price_change_df <- coke_price_change %>% 
  dplyr::bind_rows(pepsi_price_change) %>% 
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = diff.xts)

# make a scatter plot of Coke versus Pepsi price changes over a year
# as shown in figure 6.1
ggplot(price_change_df, aes(COKE, PEP)) + 
  geom_point(shape = 1, size = 3) + 
  theme_bw() + 
  labs(x = "Coke price changes", y = "Pepsi price changes", 
       title = "Pepsi versus Coke price changes")
```

## Why pair trading?

This is how we can create a **stationary** time series for an artificial stock, called **spread**. 
Importantly, this artificial stock is a great candidate for mean-reverting, and its price can be represented as 

$$\text{spread price(t)} = \frac{x(t) A(t) + y(t) B(t)}{x(t) + y(t)}$$
$A(t)$ and $B(t)$ represent the stock price for stock A and B at time $t$. 
$x(t)$ and $y(t)$ represent the shares that one has for stock A and B at time $t$

We can see that essentially the spread price is a weighted average of stock price A and B. 

$$
\begin{align}
w_A &= \frac{x(t)}{x(t) + y(t)} \\
w_B &= \frac{y(t)}{x(t) + y(t)} \\
\text{spread price(t)} &= w_A A(t) + w_B B(t)
\end{align}
$$

Since the time series $\text{spread price(t)}$ is stationary, in other words the spread price does not change over time, this indicates $\text{spread price(t + 1)} \approx \text{spread price(t)}$. 
Meanwhile $\text{spread price(t + 1)} = w_A A(t+1) + w_B B(t+1)$. 
This means

$$
\begin{align}
w_A A(t+1) + w_B B(t+1) &\approx w_A A(t) + w_B B(t) \\
w_A [A(t+1) - A(t)] &\approx -w_B [B(t+1) - B(t)] \\
w_A \times \text{price change A} &\approx -w_B \times \text{price change B} \\
\text{price change A} &\approx -\frac{w_B}{w_A} \times \text{price change B} \\
\text{price change A} &= -\frac{w_B}{w_A} \times \text{price change B} + \epsilon
\end{align}
$$
$\epsilon$ is the random error which is normally distributed.
The $-\frac{w_B}{w_A}$ is the regression coefficient $\beta$. 
$\frac{w_B}{w_A}$ is also called **hedge ratio**. 

## How does pair traiding work?

1. estimate the spread beta $\beta$ based on the price change data. 
2. calculate the proportion of stock A based on the $\beta$.  
3. construct the spread based on the weighted average of stock price of A and B

$$
\begin{align}
\beta &= -\frac{w_B}{w_A} \\
-\frac{w_B}{w_A} &= 1 - \frac{1}{w_A} \\
\beta &= 1 - \frac{1}{w_A} \\
w_A &= \frac{1}{1 - \beta} \\
w_A &= \frac{x(t)}{x(t) + y(t)} \\
y(t) &= \frac{x(t)}{w_A} - x(t) \\
\Delta x &= x(t) - x(t-1) \\
\Delta y &= y(t) - y(t-1) \\ 
&= \frac{x(t)}{w_A} - x(t) - [\frac{x(t-1)}{w_A} - x(t-1)] \\
&= \frac{\Delta x}{w_A} - \Delta x
\end{align}
$$

$\Delta x$ ($\Delta y$) is the number of shares we need to tread (buy if it is positive, or sell if it is negative). 
Since $0 \le w_A \le 1$, $\Delta y$ must be $< 0$ if $\Delta x > 0$. 
This explains why in the pair trading strategy when one purchases stock A and he needs to sell stock B. 

## Example spread between Coke and Pepsi

```{r estimation-spread-beta}
calculate_spread_beta <- function(price_change_a, price_change_b) {
  my_fit <- lm(price_change_a ~ price_change_b)
  spread_beta <- my_fit$coefficients["price_change_b"]
  return(spread_beta)
}

calculate_spread_beta(price_change_df$COKE, price_change_df$PEP)
```

```{r in-sample-spread}
# define the function to calculate the spread price per share
calculate_spread <- function(p1, p2, b) {
  # p1 = stock A price per share at time t
  # p2 = stock B price per share at time t
  # b = spread beta
  
  w1 <- 1 / (1 - b) # proportion of stock A 
  w2 <- 1 - w1 # proportion of stock B
  spread_price <- w1 * p1 + w2 * p2
  return(spread_price)
}

# to match the figure 7.5 in the book "How to Build Your Own Algorithmic Trading Business"
# the start and end date needs to be changed
# in addition, in the same book the symbol for Coke is KO
coke <- tq_get("KO", get = "stock.prices", 
               from = "1977-01-01", to = "2008-12-31")
pepsi <- tq_get("PEP", get = "stock.prices", 
                from = "1977-01-01", to = "2008-12-31")

coke_price_change <- calculate_price_change(coke) 
pepsi_price_change <- calculate_price_change(pepsi)

# make a data frame with daily price change for Coke and Pepsi
price_change_df <- coke_price_change %>% 
  dplyr::bind_rows(pepsi_price_change) %>% 
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = diff.xts)

# based on the price change data, calculate the spread beta
spread_beta <- calculate_spread_beta(price_change_df$KO, price_change_df$PEP)

# construct a spread for Coke and Pepsi pair trading
# that is similar to the figure 7.5 in the book "How to Build Your Own Algorithmic Trading Business"
coke_pepsi_spread <- calculate_spread(p1 = coke$adjusted, p2 = pepsi$adjusted, b = spread_beta)

tibble::tibble(date = coke$date, spread = coke_pepsi_spread) %>% 
  ggplot(aes(date, spread)) + 
  geom_line() + 
  theme_bw() + 
  labs(x = NULL, y = "Spread price (in $)")
```

As we can see the spread price series is not stationary. 
