---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyquant)
library(magrittr)
library(ggplot2)
```

```{r coke-pepsi}
# download daily prices for Coke and Pepsi stock from Yahoo Finance
coke <- tq_get("COKE", get = "stock.prices", 
               from = "2013-01-01", to = "2014-01-01")
pepsi <- tq_get("PEP", get = "stock.prices", 
                from = "2013-01-01", to = "2014-01-01")

# it is assumed that if Coke goes up on a particular day, 
# then so should Pepsi. 
# as a result, we focus on price changes of these two stocks

calculate_price_change <- function(stock_price_df) {
  price_change_df <- stock_price_df %>% 
    tq_mutate(select = adjusted, mutate_fun = diff.xts) %>% 
    dplyr::select(symbol, date, diff.xts) %>% 
    dplyr::filter(!is.na(diff.xts)) # the first entry naturally contains a missing value (NA) because there is no previous price
  
  return(price_change_df)
}


coke_price_change <- calculate_price_change(coke) 
pepsi_price_change <- calculate_price_change(pepsi)

# make a data frame with daily price change for Coke and Pepsi
price_change_df <- coke_price_change %>% 
  dplyr::bind_rows(pepsi_price_change) %>% 
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = diff.xts)

# make a scatter plot of Coke versus Pepsi price changes over a year
# as shown in figure 6.1
ggplot(price_change_df, aes(COKE, PEP)) + 
  geom_point(shape = 1, size = 3) + 
  theme_bw() + 
  labs(x = "Coke price changes", y = "Pepsi price changes", 
       title = "Pepsi versus Coke price changes")
```

## Why pair trading?

This is how we can create a **stationary** time series for a artificial stock, called **spread**. 
The price of this artificial stock can be represented as 

$$\text{spread price(t)} = \frac{x A(t) + y B(t)}{x + y}$$
$A(t)$ ad $B(t)$ represent the stock price for stock A and B at the time $t$. 

## Why one needs to purchase stock A and sell stock B in pair trading?

Since the time series $\text{spread price(t)}$ is stationary, in other words the spread price does not change over time, this indicates $\text{spread price(t + 1)} \approx \text{spread price(t)}$. 
Meanwhile $\text{spread price(t + 1)} = \frac{x A(t + 1) + y B(t + 1)}{x + y}$. 
This means $x A(t + 1) + y B(t + 1) \approx x A(t) + y B(t)$. 
If we rearrange the equation we have 
$$
\begin{align}
x A(t + 1) - x A(t) &\approx  y B(t) - y B(t + 1) \\
x[A(t + 1) - A(t)] &\approx -y[B(t + 1) - B(t)]
\end{align}
$$
There is a negative sign in front of $y$, meaning selling $y$ shares of stock B. 
In contrast, stock A needs to be purchased at $x$ shares. 

## How many shares stock A need to be purchased and how many shares stock B need to be sold?

Based on the equation 
$$x[A(t + 1) - A(t)] \approx -y[B(t + 1) - B(t)]$$
we can rewrite the $\approx$ from the perspective of linear regression. 
$$
\begin{align}
x \times \text{price change A} &\approx -y \times \text{price change B} \\
\text{price change A} &\approx \frac{-y}{x} \times \text{price change B} \\
\text{price change A} &= \frac{-y}{x} \times \text{price change B} + \epsilon
\end{align}
$$
$\epsilon$ is the random error which is normally distributed.
The $\frac{-y}{x}$ is the regression coefficient $\beta$ and estimated based on the price change data of stock A and stock B. 
This provides another reason why we need to work with the price change of Coke and Pepsi. 

Later, when it is decided to buy x shares stock A, then $y = \beta \times x$ shares stock B needs to be sold. 

```{r estimation-spread-beta}
calculate_spread_beta <- function(price_change_a, price_change_b) {
  my_fit <- lm(price_change_a ~ price_change_b)
  spread_beta <- my_fit$coefficients["price_change_b"]
  return(spread_beta)
}

calculate_spread_beta(price_change_df$COKE, price_change_df$PEP)
```

```{r in-sample-spread}
# define the function to calculate the spread price per share
calculate_spread <- function(p1, p2, b) {
  # p1 = stock A price per share at time t
  # p2 = stock B price per share at time t
  # b = spread beta
  
  s1 <- 1 # shares of stock A to buy
  s2 <- s1 * b # shares of stock B to sell
  s_tot <- s1 + s2 # total shares of this artificial stock called spread
  spread_price <- (s1 * p1 + s2 * p2) / s_tot
  return(spread_price)
}

# to match the figure 7.5 in the book "How to Build Your Own Algorithmic Trading Business"
# the start and end date needs to be changed
# in addition, in the same book the symbol for Coke is KO
coke <- tq_get("KO", get = "stock.prices", 
               from = "1977-01-01", to = "2000-12-31")
pepsi <- tq_get("PEP", get = "stock.prices", 
                from = "1977-01-01", to = "2000-12-31")

coke_price_change <- calculate_price_change(coke) 
pepsi_price_change <- calculate_price_change(pepsi)

# make a data frame with daily price change for Coke and Pepsi
price_change_df <- coke_price_change %>% 
  dplyr::bind_rows(pepsi_price_change) %>% 
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = diff.xts)

# based on the price change data, calculate the spread beta
spread_beta <- calculate_spread_beta(price_change_df$KO, price_change_df$PEP)

# construct a spread for Coke and Pepsi pair trading
# that is similar to the figure 7.5 in the book "How to Build Your Own Algorithmic Trading Business"
coke_pepsi_spread <- calculate_spread(p1 = coke$adjusted, p2 = pepsi$adjusted, b = spread_beta)

tibble::tibble(date = coke$date, spread = coke_pepsi_spread) %>% 
  ggplot(aes(date, spread)) + 
  geom_line() + 
  theme_bw() + 
  labs(x = NULL, y = "Spread (in $)")
```

This means if 1 share Coke is purchased, `r spread_beta` share Pepsi needs to be sold. 
